Evaluating Cloud-Optimized HDF5 for NASA's ICESat-2 Mission

### 2024-09-05 _tex/paper.tex
Built site for gh-pages

Because of the historical complexity of operations with the HDF5 format,
there has been a necessity to make the library thread-safe and similarly
to what happens in the Python language, the simplest mechanism to
implement this is to have a global API lock. This global lock is not as
big of an issue when we read data from local disk but it becomes a major
bottleneck when we read data over the network because each read is
sequential and latency in the cloud is exponentially bigger than local
access.

@@

\subsubsection{Background and data
selection}\label{background-and-data-selection}

As a result of community feedback and ``hack weeks'' organized by NSIDC
and UW eScience Institute in 2023, NSIDC started the Cloud Optimized
Format Investigation (COFI) project to improve access to HDF5 from the
ICESat-2 mission. A spaceborne lidar that retrieves surface topography
of the Earth's ice sheets, land and (oceans Neumann et al., 2019).
Because of its complexity, large size and importance for cryospheric
studies we targeted the ATL03 dataset. ATL03 core data are geolocated
photon heights from the ICESat-2 ATLAS instrument. Each file contains
1003 geophysical variables in 6 data groups. Although our research was
focused on this dataset, most of our findings are applicable to any
dataset stored in HDF5 and NetCDF4.

@@

We tested access times to original and cloud-optimized small (1 GB),
medium (2 GB) and large (7 GB) HDF5 ATL03 files {[}list files tested{]}
stored in AWS S3 buckets in region us-west-2, the region hosting NASA's
Earthdata Cloud archives. Files were accessed using Python tools
commonly used by Earth scientists: h5py and Xarray. h5py is a Python
wrapper around the HDF5 C API. xarray\footnote{\texttt{h5py} is a
  dependency of Xarray} is a widely used Python package for working with
n-dimensional data. We also tested access times using h5coro, a python
package optimized for reading HDF5 files from S3 buckets and kerchunk, a
tool that creates an efficient lookup table for file chunks to allow
performant partial reads of files.

HDF5 ATL03 files were originally cloud optimized by ``repacking'' them,
using a relatively new feature in the HDF5 C API called ``paged
aggregation''. Page aggregation does 2 things: it collects file-level
metadata from datasets and stores it on dedicated metadata blocks in the
file; and it forces the library to write data using fixed-size blocks.
Aggregation allows client libraries to read file metadata with only a
few requests and uses the page size used in the aggregation as the
minimal request size, overriding the 1 request per chunk behavior.

@@

@@

\caption{\label{fig-3}Benchmarks show that cloud optimizing ATL03 files
improved access times at least an order of magnitude when used with
aligned I/O patterns, this is telling the library about the cloud
optimization and page size.}

@@

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

@@

@@

>>>>

Because of the historical complexity of operations with the HDF5
format(The HDF Group, n.d.), there has been a necessity to make the
library thread-safe and similarly to what happens in the Python
language, the simplest mechanism to implement this is to have a global
API lock. This global lock is not as big of an issue when we read data
from local disk but it becomes a major bottleneck when we read data over
the network because each read is sequential and latency in the cloud is
exponentially bigger than local access (MDN, 2024) (Scott, 2020).

@@

\subsubsection{\texorpdfstring{\textbf{Background and data
selection}}{Background and data selection}}\label{background-and-data-selection}

As a result of community feedback and ``hack weeks'' organized by NSIDC
and UW eScience Institute in 2023(ICESAT-2 HackWeek, 2023), NSIDC
started the Cloud Optimized Format Investigation (COFI) project to
improve access to HDF5 from the ICESat-2 mission. A spaceborne lidar
that retrieves surface topography of the Earth's ice sheets, land and
oceans (Neumann et al., 2019). Because of its complexity, large size and
importance for cryospheric studies we targeted the ATL03 dataset. ATL03
core data are geolocated photon heights from the ICESat-2 ATLAS
instrument. Each file contains 1003 geophysical variables in 6 data
groups. Although our research was focused on this dataset, most of our
findings are applicable to any dataset stored in HDF5 and NetCDF4.

@@

We tested access times to original and different configurations of
cloud-optimized HDF5 ATL03 files {[}list files tested{]} stored in AWS
S3 buckets in region us-west-2, the region hosting NASA's Earthdata
Cloud archives. Files were accessed using Python tools commonly used by
Earth scientists: h5py and Xarray(Hoyer \& Hamman, 2017). h5py is a
Python wrapper around the HDF5 C API. xarray\footnote{\texttt{h5py} is a
  dependency of Xarray} is a widely used Python package for working with
n-dimensional data. We also tested access times using h5coro, a python
package optimized for reading HDF5 files from S3 buckets and kerchunk, a
tool that creates an efficient lookup table for file chunks to allow
performant partial reads of files.

The test files were originally cloud optimized by ``repacking'' them,
using a relatively new feature in the HDF5 C API called ``paged
aggregation''. Page aggregation does 2 things: it collects file-level
metadata from datasets and stores it on dedicated metadata blocks in the
file; and it forces the library to write both data and metadata using
these fixed-size pages. Aggregation allows client libraries to read file
metadata with only a few requests and uses the page size used in the
aggregation as the minimal request size, overriding the 1 request per
chunk behavior.

@@

As we can see in Figure~\ref{fig-2}, when we cloud optimize a file using
paged-aggregation there are some considerations and behavior that we had
to take into account. The first thing to observe is that page
aggregation will --as we mentioned-- consolidate the file-evel metadata
at the front of the file and will add information in the so-called
superblock\footnote{The HDF5 superblock is a crucial component of the
  HDF5 file format, acting as the starting point for accessing all data
  within the file. It stores important metadata such as the version of
  the file format, pointers to the root group, and addresses for
  locating different file components} The next thing to notice is that
page size us uses across the board for metadata and data as of October
2024 and version 1.14 of the HDF5 library, page size cannot dynamically
adjust to the total metadata size.

@@

\caption{\label{fig-3}shows how file-level metadata and data packing
inside aggregated pages leave unused space that can potentially increase
the file size in a considerable way.}

This one page size for all approach simplifies how the HDF5 API reads
the file (if instructucted) but it also brings unused page space and
chunk over-reads. In the case of the ICESat-2 dataset (ATL03) the data
itself has been partitioned and each granule represents a segment in the
satellite orbit and within the file the most relevant dataset is chunked
using 10,000 items per chunk, with data being float-32 and using a fast
compression value, the resulting chunk size is on average under 40KB,
which is really small for an HTTP request, especially when we have to
read them sequentially. Because of these considerations, we opted for
testing different page sizes, and increase chunk size. The following
table describes the different configurations used in our tests.

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1589}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.3444}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1457}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0993}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.0728}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 12\tabcolsep) * \real{0.1060}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
prefix
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
description
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\% file size increase
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textasciitilde km per chunk
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
shape
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
page size
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
avg\_chunk\_size
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
original & original file from ATL03 v6 (1gb and 7gb) & 0 & 1.5km &
(10000,) & N/A & 35kb \\
original-kerchunk & kerchunk sidecar of the original file & N/A & 1.5km
& (10000,) & N/A & 35kb \\
page-only-4mb & paged-aggregated file with 4mb per page &
\textasciitilde1\% & 1.5km & (10000,) & 4MB & 35kb \\
page-only-8mb & paged-aggregated file with 4mb per pag8 &
\textasciitilde1\% & 1.5km & (10000,) & 8MB & 35kb \\
rechunked-4mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 4MB & 400kb \\
rechunked-8mb & page-aggregated and bigger chunk sizes &
\textasciitilde1\% & 10km & (100000,) & 8MB & 400kb \\
rechunked-8mb-kerchunk & kerchunk sidecar of the last paged-aggregated
file & N/A & 10km & (10000,) & 8MB & 400kb \\
\end{longtable}

This table represents the different configurations we used for our tests
in 2 file sizes. It's worth noticing that we encountered a few outlier
cases where compression and chunk sizes led page aggregation to an
increase in file size of approximately 10\% which was above the desired
value for NSIDC (5\% max) We tested these files using the most common
libraries to handle HDF5 and 2 different I/O drivers that support remote
access to AWS S3, fsspec and the native S3. The results of our testing
is explained in the next section and the code to reproduce the results
is in the attached notebooks.

\includegraphics{figures/figure-5.png}

\caption{\label{fig-5}Benchmarks show that cloud optimizing ATL03 files
improved access times at least an order of magnitude when used with
aligned I/O patterns, this is telling the library about the cloud
optimization and page size.}

@@

\section{References}\label{references}

@@

\bibitem[\citeproctext]{ref-Hoyer2017-su}
Hoyer, S., \& Hamman, J. (2017). Xarray: {N-D} labeled arrays and
datasets in python. \emph{J. Open Res. Softw.}, \emph{5}(1), 10.

\bibitem[\citeproctext]{ref-h5cloud2023}
ICESAT-2 HackWeek, H. C. (2023). h5cloud: Tools for cloud-based analysis
of HDF5 data (Version v1.0.0). Retrieved from
\url{https://github.com/ICESAT-2HackWeek/h5cloud}

\bibitem[\citeproctext]{ref-Mozilla-latency-2024}
MDN, M. (2024, May). Understanding latency. Retrieved from
\url{https://developer.mozilla.org/en-US/docs/Web/Performance/Understanding_latency}

@@

\bibitem[\citeproctext]{ref-scott-2020}
Scott, C. (2020). Numbers every programmer should know. Retrieved from
\url{https://colin-scott.github.io/personal_website/research/interactive_latency.html}

\bibitem[\citeproctext]{ref-The_HDF_Group_Hierarchical_Data_Format}
The HDF Group. (n.d.). {Hierarchical Data Format, version 5}. Retrieved
from \url{https://github.com/HDFGroup/hdf5}
